---
title: "Quarto slides stuff"
format: revealjs
---


https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/
https://info5940.infosci.cornell.edu/slides/publishing-reproducible-documents/#/themesappearance

# Q slides columns

## Two cols slide
:::: {.columns}

::: {.column width="40%"}
contents...
:::
  
::: {.column width="60%"}
contents...
:::
  
::::

## Q slides incr list
::: {.incremental}
- ELement  1
- ELement  2
:::
  
## Slide Code block ... or 


```{r code-line-numbers="4:14"}
#| output-location: fragment

 
```


## Slide Code block



## Slide Code + output (next slide)
https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/posts/05-output-location/

```{r code-line-numbers="4:14"}
#| output-location: slide

 
```

```{r}
#| echo: true
#| output-location: slide
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```
 
## Slide Code + output (fragment)
https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/posts/05-output-location/

```{r}
#| echo: true
#| output-location: fragment
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```

## Slide Code + output (column)
https://mine-cetinkaya-rundel.github.io/quarto-tip-a-day/posts/05-output-location/

```{r}
#| echo: true
#| output-location: column
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```  

## Slide Tabsets 
::: {.panel-tabset}

### Tab A

Content for `Tab A`

### Tab B

Content for `Tab B`

:::

## Slide Tabsets Echo + output

::: {.panel-tabset}


### Tab B
```{r}
#| echo: true
#| output-location: slide
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```

### Tab A
```{r}
library(ggplot2)
ggplot(airquality, aes(Temp, Ozone)) + 
  geom_point() + 
  geom_smooth(method = "loess")
```
 
:::

## Slide Title {.smaller}

## Slide Title  scrollable

Slide Title very very very very very very very very very very very very very very very very very very {.scrollable}


## Slide with speaker notes

Slide content

::: {.notes}
Speaker notes go here.
:::

## Slide with Aside / Footnote

- Green ^[A footnote]
- Brown
- Purple

::: aside
Some additional commentary of more peripheral interest.
:::  
  
  
  
  
  