---
title: "Installation and setup"
# general Output Options
execute:     
  eval: false
  echo: true     #  include source code in output
  warning: false  #  include warning code in output
  error: false    #  include error code in output
  output: false   # include output code in output (CHG in BLOCKS)
  # include: false   # R still runs but code and results DON"T appear in output  
  # cache: false
toc: true
format:
  html:
    code-fold: false # redundant IF echo false 
    toc-depth: 2
    toc_float: true
    toc-location: left
    toc-title: ""
---

::: {.callout-caution collapse="false"}
Website construction in progress...
:::

This workshop showcases introductory bio statistics concepts using the open source (and free!) programming language {{< fa brands r-project >}}. Each session of the workshop features exercises that will help you learn by doing. 

We will quickly go over this at our first Practice session, but you are expected to have the required software installed on your machine **before** the workshop.

Below is a step by step process that should guide you through the needed installation steps.

# Install `r fontawesome::fa("r-project", fill = "#5F85A7")`

<!-- `r fontawesome::fa("fab fa-windows", fill = "steelblue")` -->

<!-- `r fontawesome::fa("fab fa-linux", fill = "steelblue")` -->

<!-- `r fontawesome::fa("fab fa-apple", fill = "steelblue")` -->

**R** is available for free for Windows [{{< fa brands windows >}}]{style="color:#d8717b"}, GNU/Linux [{{< fa brands linux >}}]{style="color:#d8717b"}, and macOS [{{< fa brands apple >}}]{style="color:#d8717b"}.

-   To install **R**, you can go to this [**link**](https://cloud.r-project.org/). The latest available release is **R 4.3.3 "Angel Food Cake" released on 2024-02/29**, but any (fairly recent) version will do.

If you have previously installed R on your machine, you can check which version you are running by executing this command in `R`:

```{r}
#| eval: false
#| echo: true

# From the R console
base::R.version.string
    # (This is the version on my own machine)
    # [1] "R version 4.2.2 (2022-10-31)"
```

...or by executing this command in your `CLI` (Command Line Interface):

```{bash}
#| eval: false
#| echo: true

# From Terminal/Powershell/bash
R --version
```

# Install RStudio IDE

While not strictly required, it is highly recommended that you also install RStudio to facilitate your work. **RStudio Desktop** is an Integrated Development Editor (IDE), basically a graphical interface wrapping and interfacing R (which needs to be installed first).

::: {.callout-tip collapse="true"}
R, which is a command line driven program, can be executed via its native interface (**R GUI**), as well as from many other code editors, like **VS Code**, **Sublime Text**, **Jupyter Notebook**, etc. **RStudio** remains the most widely used by beginners and advanced programmers alike, because of its intuitive and integrated interface.
:::

- To install **RStudio** you can go to this [**link**](https://posit.co/download/rstudio-desktop/).
- The free **RStudio DEsktop** version contains everything you need.
- Again, you don't need to have the latest RStudio version, but I recommend **v2022.07.1** or later (it also includes the installation of Quarto)

::: {.callout-tip collapse="true"}
[Quarto](https://quarto.org/) is a multi-language, next-generation version of R Markdown from Posit (former RSTUDIO company). While for R users it is not much different form `Rmarkdown` (you can author files with extension .qmd which are very similar), its key feature is that is It is available also for Python, Julia, and Observable languages. 

+ Quarto is not an R package so it has to be downloaded separately. However Rstudio v2022.07.1 or later come with Quarto included

+ The [www.R4biostats.com](www.R4biostats.com) is actually a quarto website 😊
 
:::

![RStudio Pane Layout [Source: Posit's RStudio User Guide](https://docs.posit.co/ide/user/ide/guide/ui/ui-panes.html)](images/rstudio-panes-labeled.jpeg) 

# Install R packages from the CRAN

An **R *package*** is a shareable bundle of functions. Besides the basic built-in functions already contained in the program (i.e. the `base` package), many useful R functions come in free libraries of code (or *packages*) written by R's user community.

-   **CRAN** - the Comprehensive R Archive Network - is the general package repository for R: [https://cran.r-project.org/](https://cran.r-project.org/)

(We only use packages from CRAN for our purposes) 

-   **Bioconductor** - a  package repository geared towards biostatistics: [https://www.bioconductor.org/](https://www.bioconductor.org/)

-   **Github** - a repository where you can find developer’s version of a package: [https://github.com/](https://github.com/)


# R packages that will be required for the workshop

Below are the R packages, that we will need for the workshop's practical sessions. Ideally, you can pre-install them... it will also serve as a test to check whether the R version you have is compatible. 


```{r}
#| eval: false
#| echo: true

# Installing (only the 1st time you use them)
pkg_list <- c("here", "fs", "janitor", "skimr",
              "dplyr", "tidyr","forcats", "ggplot2","ggridges")

install.packages(pkg_list)

# Loading a package (at the beginning of every R session) 
library(here)     # A Simpler Way to Find Your Files      
library(fs)       # Cross-Platform File System Operations     
library(janitor)  # Simple Tools for Examining and Cleaning Data   
library(skimr)    # Compact and Flexible Summaries of Data     
library(dplyr)    # {tidyverse} A Grammar of Data Manipulation     
library(tidyr)    # {tidyverse} Tools tocreate tidy data 
library(forcats)  # {tidyverse} Tools for Categorical Var.(Factors)   
library(ggplot2)  # {tidyverse} Tools for Data Visualizations    
library(ggridges) # Ridgeline Plots in 'ggplot2' (density functions)
```


# R coding tips and tricks 

## Useful keyboard shortcuts in RStudio

```{r}
#| eval: true
#| echo: false
#| output: true

library(tibble)
library(flextable)

key_short <- tibble::tribble (
  ~Description, 	             ~'Windows & Linux',     ~'Mac', 
  "Insert code section", 	     "Ctrl+Shift+R",       	"Shift+Command+R  or Ctrl+Shift+R",
  "Run current line/selection", "Ctrl+Enter",     "Command+Return",
  "Reindent lines", 	             "Ctrl+I", 	"Command+I",
  
  "Insert <- operator", 	"Alt+-", 	"Option+-",
  "Insert %>% operator", 	    "Ctrl+Shift+M", 	"Shift+Command+M",
  
  "Code completion",          "Tab", "Tab", 
  "File path completion (console)",    "\"+Tab", "\"+Tab"
)

 	
key_short_ft <- key_short %>% 
  # adorn_percentages(  denominator = 'row', na.rm = TRUE) %>%
  flextable() %>% 
 #   flextable::set_header_labels(key_stakes_types = "Type of agencies most frequently involved", N = "N. of cases") %>% 
  bold( bold = TRUE,   part = "header") %>% 
  color( color = "white",   part = "header") %>% 
  bg( bg = "#d8717b",   part = "header") %>% 
  bg( bg = "#F2F2F2", part = "body", source = j) %>% 
  fontsize( i = NULL, j = NULL, size = 11, part = "body") %>% 
  # bg(bg =bordeaux_light, i = ~  `Type of agencies most frequently involved` %in%  c ( "more than 1 agency involved" ), part = "body") %>% 
  fontsize(size = 14, part = "header") %>% 
  fontsize(size = 12, part = "body") %>% 
  autofit() 


key_short_ft
```

Here you find the complete list of [RStudio Keyboard Shortcuts](https://docs.posit.co/ide/user/ide/reference/shortcuts.html)
<!-- ## Install [Quarto] -->

<!--  - Install latest version of [Quarto](https://quarto.org/docs/get-started/) (v1.0.36 or greater) -->

<!-- - After installing Quarto, open the terminal in RStudio and: -->

<!--   - See [this guide](https://support.rstudio.com/hc/en-us/articles/115010737148-Using-the-RStudio-Terminal-in-the-RStudio-IDE) for details on using the terminal -->

<!--   - Install/Update TinyTeX for PDFs with `quarto install tool tinytex` -->

<!--   - Install/Update Chromium with `quarto install tool chromium` -->


