---
title: "Acknowledgments and use of the course materials"
# general Output Options
execute:     
  eval: true
  echo: false     #  include source code in output?
  warning: false  #  include warning code in output?
  error: false    #  include error code in output?
  output: true   # include output code in output (CHG in BLOCKS)?
  # include: false   # R still runs but code and results DON"T appear in output  
  # cache: false
toc: true
format:
  html:
    code-fold: false # redundant IF echo false 
    toc-depth: 2
    toc_float: true
    toc-location: left
    toc-title: ""
    # template: theme/template.tex  
bibliography: bib/R4biostats.bib
csl: bib/apa-cv.csl # Very important for complete citations!!!!! 
## --------- NO END REFERENCES .... other options are redundant!!!!!!!!!!!!!!!!!!!!!!!!!!!!!!
suppress-bibliography: true
# biblio-style: apa # authoryear-icomp # abbrv # alpha #apalike # plain # unsrt #siam
#reference-section-title: References
#link-citations: false
# filters:
#   - section-bibliographies
---

::: {.callout-caution collapse="false"}
Website construction in progress...
:::

```{r}
if (!require(pacman)) {install.packages(pacman)} 
library(pacman)

# pacman::p_ver("bib2df")
# pacman::p_exists("bib2df")  #TRUE
# pacman::p_ver_diff("bib2df")
# p_load_gh("ropensci/bib2df") # otherwise warns `as_data_frame() was deprecated in tibble 2.0.0.`

#remotes::install_github("ropensci/bib2df")
p_load(RefManageR,
       #  bib2df,
       bibtex,
       janitor,here, paint,
       glue,
       tidyverse, # lubridate,
       htmltools,
       # distilltools,
       bib2df
)

```

```{r}
# - Load bibliography (exported from Zotero)
bib_path <- here::here ("bib","R4biostats.bib")

# - as list  (NEEDED for functions later)
mybib <- RefManageR::ReadBib(bib_path, check = FALSE )  #bibtex::do_read_bib(file = bib_path, encoding = "unknown", srcfile)

# - as dataframe
mybib_df_raw <- bib2df::bib2df(bib_path)  

# - Format dataframe version of bibliography
mybib_df <-
  mybib_df_raw %>%
  clean_names() %>%
  select(key = bibtexkey, 
         date, 
         author, 
         title, 
         category,# THIS MY Zotero's 
         type, # THIS MY OWN "Extra: tex.type: ..." in zotero 
         url, 
         file  ) %>%
  mutate(
    # date = parse_date_time(date, "ymd",
    # 							  #select_formats = .select_formats
    # 							  drop = FALSE),
    yyyy_mm = str_sub(date, start = 1, end = 7), # I have some irregularities in format  
    date_parsed = parse_date_time(yyyy_mm,  "ym", truncated = 2 ),
    year = year(date_parsed), 
    month = month(date_parsed)) %>%
  arrange(desc(year), key) %>% 
  dplyr::relocate(c("yyyy_mm" ,    "year" ,        "month"  ), .after =date)

   
# # Set RefManageR options (needed to avoid the "[1]")
# oldopts <- RefManageR::BibOptions(
# 	style = "text",
# 	bib.style = "authoryear", 
# 	no.print.fields = c("doi", "url", "urldate", "eprint", "eprinttype"),
# 	max.names = 50, # maximum number of names to display before using “et al.”
# 	dashed = FALSE, 
# 	sorting = "ydnt"
# )

# # - Load additional metadata
# mybib_meta <- tibble::tribble( 
# 	~"url_pub", ~"url_pdf",~ "url_rg", ~"url_repo", ~"url_arch", ~"url_osf", ~"url_other", ~"url_blog" ) 
# mybib_meta[1:nrow(mybib_df), ] <- as.character(NA)
# 
# # bind 
# mybib_df2 <- cbind(mybib_df, mybib_meta)
# names(mybib_df2)
# rm(mybib, mybib_df, mybib_df_raw, mybib_meta)
```

 
## Acknowledgements

## Licensing and use of the course materials

something along the lines of [MADA course](https://andreashandel.github.io/MADAcourse/)   


## List of great resources 

### Section Article

```{r article}
#| include: false

mybib_df[mybib_df$type == "article",  1] 
# 
# print(mybib[type = "article"], 
#       .opts = list(bib.style = "authoryear", 
#                    check.entries = FALSE))

```

 + @wongsaengsak_significance_2019-1  
 + @kruschke_bayesian_2018     
 + @mayne_role_1957    

### Section course

```{r course}
#| include: false

mybib_df[mybib_df$type == "course",  1] 

# print(mybib[type = "course"], 
#       .opts = list(bib.style = "authoryear", 
#                    check.entries = FALSE))

```

+ @applied_epi_team_applied_2024                              
+ @bobbitt_statology_2024                                     
+ @csunedu_httpwwwcsuneduan73773slidesclass10f09pdf_2024      
+ @cuemath_z-test_2024                                        
+ @datatab_team_datatab_2024                                  
+ @ginolhac_biostatistics_2024                                
+ @httpsstatisticsbyjimcom_statisticsbyjimcom_2024            
+ @matson_icy0006_2024                                        
+ @rafalab_dsbook-part-1_2024                                 
+ @selby_analytical_2024                                      
+ @towardsdatascience_5-ways--increase-statistical-power-_2024
+ @tuos-bio-data-skills_last_scientific-process_2024          
+ @childs_introductory_2022                                   
+ @edanz-learning-team_frequentist_2022                       
+ @shahbaba_biostatistics_2012    

### Section YT course

```{r YT_course}
#| include: false

mybib_df[mybib_df$type == "YT course",  1] 
# print(mybib[type = "YT course"], 
#       .opts = list(bib.style = "authoryear", 
#                    check.entries = FALSE))
``` 

+ @very_normal_biostatistics_2024

### Section R

```{r R}
#| include: false

mybib_df[mybib_df$type == "R",  1] 

# print(mybib[type = "R"], 
#       .opts = list(bib.style = "authoryear", 
#                    check.entries = FALSE))
``` 
+ @applied_epi_team_r_2024           
+ @codoni_course_2024                
+ @dplyr_vignette_programming_2024   
+ @rstudio_base-r-cheat-sheetpdf_2024
+ @coolbutuseless_anonymous_2019     
+ @wickham_tidy_2014                 
+ @logan_biostatistical_2010  

### Section Rdataset 

```{r dataset}
#| include: false

mybib_df[mybib_df$type == "dataset",  1] 
# print(mybib[type = "dataset"],
#        .opts = list(bib.style = "authoryear", 
#                     check.entries = FALSE))
```

+ @chicco_machine_2020           
+ @wongsaengsak_significance_2019
+ @ahmad_survival_2017 
  
<!-- <!-- For HTML output -->  
<!-- ::: {.content-hidden when-format="html"} -->
<!-- ## References -->
<!-- ::: -->

